\chapter*{ABSTRACT}%
\begin{justify}
\addcontentsline{toc}{chapter}{ABSTRACT}

High-Pressure Die Casting (HPDC) is widely used in automotive manufacturing to produce structural aluminium components at high production rates. Surface defects --- particularly \textbf{Cracks} and \textbf{Cold Shuts} --- pose structural safety risks and must be reliably identified before parts leave the production line. Current end-of-line (EOL) quality control at Magna International relies on human visual inspection, which is prone to fatigue-induced inconsistency, and X-Ray analysis, which is limited to three parts per shift and is prohibitively slow for 100\% inspection coverage. This project addresses that gap by developing an automated, commercially deployable deep learning pipeline for surface defect detection.

The dataset comprises 256 high-resolution HPDC casting images (4096$\times$3000 pixels) annotated in COCO polygon format, provided by Magna International. Exploratory data analysis established that crack bounding boxes occupy as little as 0.065\% of the total image area, placing cracks firmly in the extreme small-object regime where conventional full-image detectors fail. A patch-based sliding-window pipeline was therefore designed to generate 1024$\times$1024 overlapping patches (35\% overlap) from each full image, with Shapely polygon-aware annotation clipping to preserve accurate ground-truth boundaries.

The initial model candidate was YOLOv8, but its AGPL-3.0 licence is incompatible with commercial deployment at Magna International, as it requires derivative products to be open-sourced. The team therefore adopted \textbf{RF-DETR Medium} (Apache 2.0), a Detection Transformer with a DINOv2 ViT-B backbone and multi-scale deformable attention decoder, which permits unrestricted commercial use. Seven structured experiments progressively refined the pipeline through dataset balancing (1:2 positive:negative ratio), crack-focused horizontal-flip augmentation, and patch size scaling. The final configuration achieved an EMA mAP50-95 of \textbf{0.2379} and mAP50 of \textbf{0.4647}, representing an approximately 11\% relative improvement over the baseline. Dataset engineering --- particularly increasing patch size from 768 to 1024 pixels with 35\% overlap --- proved to be the dominant performance driver.

\end{justify}
\pagenumbering{roman}

\begin{flushleft}
\vspace{1cm}
\textsl{\textbf{Keywords:} HPDC casting, RF-DETR, object detection, crack detection, deep learning, DINOv2, patch-based training, Deformable DETR, COCO mAP, Magna International.}
\end{flushleft}

\pagenumbering{roman}
