\chapter*{ACKNOWLEDGEMENT}

%\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENT}

\begin{justify}

We would like to thank our faculty and management for their professional guidance towards
the completion of the minor project work. We take this opportunity to thank Dr. Ashok
Shettar, Pro-Chancellor, Dr. P.G Tewari, Vice-Chancellor and Dr. B.S.Anami, Registrar for
their vision and support. \\ \\
We also take this opportunity to thank Dr. Meena S.M, Professor and Dean of Faculty, SoCSE,
Dr. Vijayalakshmi M, Professor, Head of CSE for having provided us direction and facilitated
for enhancement of skills and academic growth. \\ \\
We thank our guide \textbf{Neha Tarannum}, Designation, Department of Computer Science \&
Engineering for the constant guidance during interaction and reviews. \\ \\
We extend our acknowledgment to the reviewers for critical suggestions and inputs. We also
thank the project coordinator Dr. Uday Kulkarni, Mr. Sunil V. Gurlahosur and reviewers for
their suggestions during the course of completion. We express gratitude to our beloved
parents for constant encouragement and support.

\end{justify}

\begin{flushright}
\vspace{3cm}
Sushanth Tiwari -- 01FE23BCS086\\
\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent
Sinchana S M -- 01FE23BCS192\\
\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent\indent
Akshata Shet -- 01FE23BEV035\\
\end{flushright}
