\chapter{IMPLEMENTATION AND EXPERIMENTS}

\justify

This chapter describes the implementation methodology and key modules of the HPDC Defect Detection System, followed by the three principal experiments conducted to systematically improve detection performance. The implementation comprises five functional modules: dataset analysis, patch generation, dataset engineering, RF-DETR training, and evaluation and inference. The experimental methodology is structured to isolate and quantify the contribution of individual pipeline choices.

\section{Proposed Methodology}
\justify

The core methodology is a \textbf{patch-based sliding-window training pipeline} designed for extreme small-object detection. Since crack bounding boxes occupy less than 0.1\% of the full image area, full-image detection is infeasible. Each HPDC image is divided into overlapping 1024$\times$1024 patches using a stride of 665 px (35\% overlap), with Shapely polygon-aware annotation clipping to preserve accurate ground-truth boundaries.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=0.5cm,
block/.style={rectangle, rounded corners=4pt, draw=magnaDark,
fill=magnaLightGray, text width=7.5cm,
minimum height=1cm, align=center, font=\small},
arrow/.style={-Stealth, thick, color=magnaRed}]
\node[block, fill=magnaBlue!15, draw=magnaBlue] (input)
    {\textbf{Input:} Full image + COCO polygon annotations};
\node[block, below=0.5cm of input] (stride)
    {\textbf{Sliding Window:} Stride $= 1024 \times (1 - 0.35) = 665$\,px};
\node[block, below=0.5cm of stride] (extract)
    {\textbf{Patch Extraction:} Crop \texttt{image[y:y+H, x:x+W]}};
\node[block, below=0.5cm of extract] (shapely)
    {\textbf{Polygon Clipping (Shapely):} Intersect polygon with patch boundary;
    recompute bbox from clipped polygon};
\node[block, below=0.5cm of shapely] (balance)
    {\textbf{Balancing:} Downsample negatives to 1:2 pos:neg ratio};
\node[block, below=0.5cm of balance] (aug)
    {\textbf{Augmentation:} Horizontal flip on crack patches only};
\node[block, fill=successGreen!10, draw=successGreen, below=0.5cm of aug] (output)
    {\textbf{Output:} COCO-format JSON + PNG patches for training};
\draw[arrow] (input)   -- (stride);
\draw[arrow] (stride)  -- (extract);
\draw[arrow] (extract) -- (shapely);
\draw[arrow] (shapely) -- (balance);
\draw[arrow] (balance) -- (aug);
\draw[arrow] (aug)     -- (output);
\end{tikzpicture}
\caption{Implementation methodology flowchart: patch-based data engineering pipeline.}
\label{fig:methodology}
\end{figure}

The EMA training strategy smooths weight updates:
\begin{equation}
\boldsymbol{\theta}_{\text{EMA}} = \alpha \cdot \boldsymbol{\theta}_{\text{EMA}} + (1-\alpha) \cdot \boldsymbol{\theta}_{\text{current}}
\end{equation}
where $\alpha \approx 0.9999$. EMA weights consistently outperformed instantaneous weights across all experiments.

\section{Module Descriptions}

\subsection{Module 1: Dataset Analysis Module}

\textbf{Input:} Raw COCO JSON annotation file for the full 256-image dataset.

\textbf{Output:} Statistical summary of crack/cold shut sizes, class distribution, annotation quality report.

\textbf{Description:} This module performs exploratory data analysis (EDA) to characterise the defect dataset. It computes bounding box area distributions, identifies that crack areas occupy a median of 0.065\% of image area, and validates polygon annotation quality. The EDA confirmed that full-image training is infeasible and a patch-based strategy is required.

\begin{lstlisting}[language=Python, caption={Dataset analysis pseudocode}]
def dataset_analysis(coco_json_path):
    coco = load_coco(coco_json_path)
    image_area = 4096 * 3000  # pixels^2
    for ann in coco['annotations']:
        bbox_area = ann['bbox'][2] * ann['bbox'][3]
        relative_area = bbox_area / image_area * 100
        record(ann['category_id'], relative_area)
    # Output: median crack area ~ 0.065%, Cold Shut ~ 0.12%
    plot_distributions()
\end{lstlisting}

\subsection{Module 2: Patch Generation Module}

\textbf{Module Name:} \texttt{generate\_patches()}

\textbf{Input:} Full HPDC image (4096$\times$3000 PNG), COCO annotation JSON.

\textbf{Output:} Set of 1024$\times$1024 PNG patches + updated COCO JSON with clipped annotations.

\textbf{Description:} Enumerates sliding-window positions with stride 665 px. For each patch, clips all polygon annotations using Shapely geometry intersection and recomputes bounding boxes. Discards annotations where intersection area falls below a minimum threshold.

\begin{lstlisting}[language=Python, caption={Patch generation pseudocode}]
def generate_patches(image, annotations,
                     patch_size=1024, overlap=0.35):
    stride = int(patch_size * (1 - overlap))  # 665 px
    for y in range(0, H - patch_size + 1, stride):
        for x in range(0, W - patch_size + 1, stride):
            patch    = image.crop(x, y, patch_size, patch_size)
            patch_box = Polygon([(x,y),(x+S,y),
                                  (x+S,y+S),(x,y+S)])
            clipped = []
            for ann in annotations:
                poly  = Polygon(ann['segmentation'])
                inter = poly.intersection(patch_box)
                if inter.area > MIN_THRESHOLD:
                    bbox = inter.bounds
                    clipped.append(
                        make_coco_ann(bbox, ann['category_id']))
            save_patch(patch, clipped)
\end{lstlisting}

\subsection{Module 3: Dataset Engineering Module}

\textbf{Module Name:} \texttt{balance\_and\_augment()}

\textbf{Input:} All generated patches with annotation counts.

\textbf{Output:} Balanced and augmented patch dataset ready for training.

\textbf{Description:} Separates positive (defect-containing) and negative (background) patches. Randomly downsamples negatives to a 1:2 positive:negative ratio. Applies horizontal flip to crack-containing patches, recomputing bounding box coordinates as $x_{\text{new}} = W - x - w$.

\begin{lstlisting}[language=Python, caption={Balancing and augmentation pseudocode}]
def balance_and_augment(patches):
    pos = [p for p in patches if p.annotations]
    neg = [p for p in patches if not p.annotations]
    neg = random.sample(neg, min(len(neg), 2 * len(pos)))
    augmented = []
    for patch in pos:
        if has_crack(patch):
            flipped_img  = cv2.flip(patch.image, 1)
            flipped_anns = [flip_bbox(a, patch.width) for a in patch.annotations]
            augmented.append(Patch(flipped_img, flipped_anns))
    return pos + neg + augmented

def flip_bbox(ann, W):
    x, y, w, h = ann['bbox']
    return {'bbox': [W - x - w, y, w, h], 'category_id': ann['category_id']}
\end{lstlisting}

\subsection{Module 4: RF-DETR Training Module}

\textbf{Module Name:} \texttt{train\_rfdetr()}

\textbf{Input:} COCO-format patch dataset (train/val splits), training configuration.

\textbf{Output:} Best EMA checkpoint (\texttt{checkpoint\_best\_ema.pth}), per-epoch mAP metrics.

\textbf{Description:} Fine-tunes RF-DETR Medium with DINOv2 ViT-B backbone using Hungarian bipartite matching loss (classification + $\mathcal{L}_1$ + GIoU). EMA weights are maintained and saved when validation EMA mAP50-95 improves.

\begin{lstlisting}[language=Python, caption={RF-DETR training configuration (Experiment 7)}]
model = RFDETRMedium(pretrain_weights='rf-detr-medium.pth')
model.train(
    dataset_dir = 'RF_DETR_Medium_1024/',
    epochs      = 20,
    batch_size  = 12,
    imgsz       = 1024,
    workers     = 16,
    device      = 'cuda'
)
# EMA mAP50-95 = 0.2379, mAP50 = 0.4647 at Epoch 10
\end{lstlisting}

\subsection{Module 5: Evaluation and Inference Module}

\textbf{Module Name:} \texttt{evaluate()} / \texttt{detect()}

\textbf{Input:} Test split patches (COCO format) or full HPDC image + trained EMA checkpoint.

\textbf{Output:} COCO mAP50-95, mAP50, per-class AP (evaluation) or annotated image with bounding boxes (inference).

\textbf{Description:} Evaluation runs the EMA checkpoint against the test split using pycocotools for standard COCO metrics. Inference generates patches from a full input image, runs RF-DETR forward pass on each patch, maps bounding boxes back to full-image coordinates, and renders the annotated output.

\begin{lstlisting}[language=Python, caption={Inference pseudocode}]
def detect(image_path, checkpoint, conf_threshold=0.3):
    model = RFDETRMedium()
    model.load_state_dict(torch.load(checkpoint))
    model.eval()
    image   = load_image(image_path)
    patches = generate_patches(image, patch_size=1024, overlap=0.35)
    results = []
    for (patch, x_off, y_off) in patches:
        dets = model(patch)
        for det in dets:
            if det.confidence > conf_threshold:
                det.bbox = offset_bbox(det.bbox, x_off, y_off)
                results.append(det)
    return annotate_image(image, results)
\end{lstlisting}

\section{Experimental Methodology}

Seven structured training experiments were conducted to systematically identify the impact of each dataset engineering decision on detection performance. Experiments 1, 4, and 7 are reported here as they represent the baseline, the best-with-balancing-and-augmentation, and the overall best configuration respectively. All experiments used RF-DETR Medium with identical model architecture and training hyperparameters; only the data pipeline configuration was varied. The primary evaluation metric is EMA mAP50-95 on the validation split.

\section{Experiment 1 --- RF-DETR Medium Baseline (768px, 20\% Overlap)}

\textbf{Objective:} Establish a reference performance level for RF-DETR Medium fine-tuned on 768$\times$768 patches with no balancing or augmentation.

\textbf{Dataset configuration:}
\begin{itemize}[leftmargin=2em]
    \item Patch size: 768$\times$768 px, stride: 614 px (20\% overlap)
    \item No negative patch downsampling; no augmentation
    \item Train: 2,776 patches; Val: 684 patches; Test: 447 patches (total 3,907)
    \item Class distribution (train): 80.2\% negative patches, 12.1\% crack-only, 7.7\% mixed
\end{itemize}

\textbf{Training configuration:} Epochs: 20, batch size: 12, imgsz: 768, workers: 16, device: CUDA (RTX 6000 Ada).

\begin{table}[H]
\centering
\caption{Experiment 1 Results}
\begin{tabular}{lcc}
\toprule
\rowcolor{tableHeader}\tableheadfont Metric & \tableheadfont Regular Checkpoint & \tableheadfont EMA Checkpoint \\
\midrule
mAP50-95       & 0.2006 & \textbf{0.2144} \\
\rowcolor{tableRowAlt}mAP50          & 0.3891 & 0.3952 \\
Crack AP50-95  & 0.154  & 0.162 \\
\rowcolor{tableRowAlt}Cold Shut AP50-95 & 0.247 & 0.267 \\
\bottomrule
\end{tabular}
\end{table}

\begin{findingbox}
\textbf{Bottlenecks identified:} (1) 80.2\% of training patches are negative (background-only), creating a severe class imbalance that biases the model towards predicting no defects. (2) At 768 px with 20\% overlap (614 px stride), many crack annotations are truncated at patch boundaries, reducing effective training signal. EMA checkpoint is 7\% better than the regular checkpoint, confirming the value of EMA tracking.
\end{findingbox}

\section{Experiment 4 --- 768px Balanced and Augmented}

\textbf{Objective:} Evaluate the isolated impact of dataset balancing and crack-specific augmentation at the 768 px patch scale.

\textbf{Dataset configuration:}
\begin{itemize}[leftmargin=2em]
    \item Patch size: 768$\times$768 px, stride: 614 px (20\% overlap) --- same as Exp 1
    \item Negative patches downsampled to 1:2 positive:negative ratio
    \item Horizontal flip applied to all crack-containing patches (bboxes recomputed)
    \item Train: 1,840 patches; Val: 684 patches; Test: 447 patches (total 2,971)
\end{itemize}

\textbf{Training configuration:} Identical to Experiment 1 (epochs: 20, batch size: 12, imgsz: 768).

\begin{table}[H]
\centering
\caption{Experiment 4 Results}
\begin{tabular}{lcc}
\toprule
\rowcolor{tableHeader}\tableheadfont Metric & \tableheadfont Regular Checkpoint & \tableheadfont EMA Checkpoint \\
\midrule
mAP50-95          & 0.2077 & \textbf{0.2157} \\
\rowcolor{tableRowAlt}mAP50             & 0.3944 & 0.4018 \\
Crack AP50-95     & 0.181  & 0.190 \\
\rowcolor{tableRowAlt}Cold Shut AP50-95 & 0.234  & 0.241 \\
\bottomrule
\end{tabular}
\end{table}

\begin{findingbox}
\textbf{Key finding:} Crack AP50-95 improved from 0.154 to 0.181, a gain of $+$17.4\% relative to Experiment 1. Dataset balancing significantly reduced the background-dominated bias. Augmentation reinforced crack detection across mirror-symmetric crack orientations. However, the overall mAP gain is modest ($+$0.6\%), suggesting that the 768 px patch size itself is a limiting factor: crack objects still appear at a very small scale within each patch, and boundary truncation at 20\% overlap continues to fragment some annotations.
\end{findingbox}

\section{Experiment 7 --- 1024px Patches with 35\% Overlap (Current Best)}

\textbf{Objective:} Determine whether increasing patch resolution from 768 to 1024 px while simultaneously increasing overlap from 20\% to 35\% produces a measurable improvement in crack detection.

\textbf{Dataset configuration:}
\begin{itemize}[leftmargin=2em]
    \item Patch size: 1024$\times$1024 px, stride: 665 px (35\% overlap)
    \item No explicit balancing (positive ratio improved naturally to 31.98\% due to larger patch coverage)
    \item Horizontal flip augmentation retained on crack patches
    \item Train: 2,939 patches (2,743 annotations); Val: 750 patches; Test: 495 patches (total 4,184)
\end{itemize}

\textbf{Training configuration:} Epochs: 20, batch size: 12, imgsz: 1024, workers: 16, device: CUDA (RTX 6000 Ada).

\begin{table}[H]
\centering
\caption{Experiment 7 Results (Current Best)}
\begin{tabular}{lcc}
\toprule
\rowcolor{tableHeader}\tableheadfont Metric & \tableheadfont Regular Checkpoint & \tableheadfont EMA Checkpoint \\
\midrule
mAP50-95          & 0.2271 & \textbf{0.2379} \\
\rowcolor{tableRowAlt}mAP50             & 0.4401 & \textbf{0.4647} \\
Crack AP50-95     & 0.198  & 0.209 \\
\rowcolor{tableRowAlt}Cold Shut AP50-95 & 0.256  & 0.267 \\
Peak epoch        & ---    & 8--10 \\
\bottomrule
\end{tabular}
\end{table}

\begin{resultbox}[Experiment 7 --- Best Configuration]
\begin{tabular}{ll}
Best EMA mAP50-95:     & \textbf{0.2379} \\
Best Regular mAP50-95: & 0.2271 \\
Best mAP50:            & \textbf{0.4647} \\
Peak epoch:            & 8--10 \\
Checkpoint:            & \texttt{checkpoint\_best\_ema.pth} \\
\end{tabular}
\end{resultbox}

\section{Comparative Summary of Experiments}

\begin{table}[H]
\centering
\caption{Comparative Results: Experiments 1, 4, and 7}
\begin{tabular}{lcccccc}
\toprule
\rowcolor{tableHeader}\tableheadfont Exp & \tableheadfont Patch & \tableheadfont Overlap & \tableheadfont Balanced & \tableheadfont Aug & \tableheadfont Reg mAP50-95 & \tableheadfont EMA mAP50-95 \\
\midrule
1 (Baseline)  & 768  & 20\% & No  & None   & 0.2006 & 0.2144 \\
\rowcolor{tableRowAlt}4 (Bal+Aug)   & 768  & 20\% & Yes & H-Flip & 0.2077 & 0.2157 \\
7 (1024+35\%) & 1024 & 35\% & No  & H-Flip & \textbf{0.2271} & \textbf{0.2379} \\
\bottomrule
\end{tabular}
\end{table}

\begin{conclusionbox}
\textbf{Key findings from the experimental series:}
\begin{enumerate}[leftmargin=2em]
    \item \textbf{Patch size is the dominant variable:} Scaling from 768 to 1024 px with 35\% overlap produced a 10.3\% relative improvement in EMA mAP50-95 (0.2157 $\to$ 0.2379), the single largest gain across all experiments.
    \item \textbf{Dataset engineering outperforms augmentation alone:} The majority of improvement came from pipeline design (patch size, overlap), not from augmentation or balancing strategies.
    \item \textbf{Crack-focused augmentation is effective:} Horizontal flip applied only to crack patches gave $+$17.4\% relative Crack AP, demonstrating that class-specific augmentation is more efficient than global augmentation.
    \item \textbf{EMA checkpoint is consistently superior:} Across all three experiments, the EMA checkpoint outperformed the regular checkpoint, confirming that EMA tracking is essential for this fine-tuning task.
\end{enumerate}
\end{conclusionbox}
