\chapter{SOFTWARE REQUIREMENT SPECIFICATION}

\justify

This chapter formally specifies the functional and non-functional requirements of the HPDC Casting Surface Defect Detection System. The system ingests high-resolution HPDC casting images, applies a patch-based deep learning pipeline, and outputs bounding box detections with class labels and confidence scores for Crack and Cold Shut defects.

\section{Overview of SRS}
\justify

The Software Requirements Specification (SRS) defines what the system must do and the constraints under which it must operate. Two primary stakeholders are identified: Magna International (end user requiring production-scale defect detection) and the university research team (developers responsible for model training, evaluation, and delivery). The SRS covers Phase 1 scope: detection and localisation of surface defects using bounding boxes on 1024$\times$1024 patches extracted from full HPDC images.

\section{Requirement Specifications}
\justify

The system is decomposed into a data engineering pipeline, a model training framework, and an inference engine. The requirements below govern all three components.

\subsection{Functional Requirements}

\begin{itemize}[leftmargin=2em]
    \item \textbf{FR1 --- Image Ingestion:} The system shall accept high-resolution HPDC casting images (up to 4096$\times$3000 pixels) in PNG or JPEG format.
    \item \textbf{FR2 --- Patch Generation:} The system shall divide input images into overlapping patches using a configurable sliding-window strategy (patch size: 768 or 1024 px; overlap: 20\%--50\%).
    \item \textbf{FR3 --- Annotation Clipping:} The system shall accurately clip COCO polygon annotations to patch boundaries using Shapely polygon-aware geometry, recomputing bounding boxes from clipped polygons.
    \item \textbf{FR4 --- Dataset Balancing:} The system shall support configurable downsampling of background (negative) patches to achieve a target positive:negative ratio (default 1:2).
    \item \textbf{FR5 --- Augmentation:} The system shall apply selective data augmentation to defect-containing patches only (horizontal flip), with automatic bounding box coordinate recomputation.
    \item \textbf{FR6 --- Model Training:} The system shall fine-tune an RF-DETR Medium model on the generated patch dataset with configurable batch size, epochs, input resolution, and worker count.
    \item \textbf{FR7 --- Checkpoint Management:} The system shall maintain both a regular and an EMA (Exponential Moving Average) checkpoint, saving the best-performing EMA checkpoint based on validation mAP50-95.
    \item \textbf{FR8 --- Evaluation:} The system shall compute and report COCO-standard metrics: mAP50-95, mAP50, and per-class AP (Crack, Cold Shut) on validation and test splits.
    \item \textbf{FR9 --- Inference:} The system shall perform patch-based inference on full HPDC images, producing bounding boxes with class labels and confidence scores mapped back to full-image coordinates.
    \item \textbf{FR10 --- Licence Compliance:} The system shall use only model architectures and libraries under licences compatible with commercial deployment (Apache 2.0, MIT, or BSD).
\end{itemize}

\subsection{Use case diagrams}

\begin{figure}[H]
\centering
\begin{tikzpicture}[
    usecase/.style={draw, ellipse, minimum width=4.2cm, minimum height=0.8cm, font=\small, align=center},
]
% System boundary
\draw[dashed, rounded corners] (1.5,-3.8) rectangle (10,3.8);
\node[font=\small\bfseries] at (5.75,4.1) {HPDC Defect Detection System};

% Actor 1 - Quality Engineer
\node[font=\small, align=center] at (-0.8, 1.5) {Quality\\Engineer};
\draw (-0.8,0.7) circle (0.28cm);
\draw (-0.8,0.42) -- (-0.8,-0.2);
\draw (-0.8,-0.2) -- (-1.1,-0.55);
\draw (-0.8,-0.2) -- (-0.5,-0.55);
\draw (-0.8,0.42) -- (-1.1,0.1);
\draw (-0.8,0.42) -- (-0.5,0.1);

% Actor 2 - ML Engineer
\node[font=\small, align=center] at (-0.8,-2.5) {ML\\Engineer};
\draw (-0.8,-3.3) circle (0.28cm);
\draw (-0.8,-3.58) -- (-0.8,-4.2);
\draw (-0.8,-4.2) -- (-1.1,-4.55);
\draw (-0.8,-4.2) -- (-0.5,-4.55);
\draw (-0.8,-3.58) -- (-1.1,-3.9);
\draw (-0.8,-3.58) -- (-0.5,-3.9);

% Use cases
\node[usecase] (uc1) at (5.75, 3.0) {Ingest HPDC Image};
\node[usecase] (uc2) at (5.75, 1.5) {Generate Patches};
\node[usecase] (uc3) at (5.75, 0.0) {Train RF-DETR Model};
\node[usecase] (uc4) at (5.75,-1.5) {Evaluate Model (mAP)};
\node[usecase] (uc5) at (5.75,-3.0) {Run Defect Inference};

% Connections
\draw (-0.5, 0.7) -- (3.65, 3.0);
\draw (-0.5, 0.7) -- (3.65,-3.0);
\draw (-0.5,-3.3) -- (3.65, 1.5);
\draw (-0.5,-3.3) -- (3.65, 0.0);
\draw (-0.5,-3.3) -- (3.65,-1.5);
\end{tikzpicture}
\caption{Use Case Diagram for HPDC Defect Detection System}
\label{fig:usecase}
\end{figure}

\subsection{Use Case descriptions using scenarios, strictly as per Pressman Template}

\begin{table}[H]
\centering
\caption{Use Case UC-05: Run Defect Detection Inference}
\begin{tabular}{p{3.5cm} p{9cm}}
\toprule
\rowcolor{tableHeader}\tableheadfont Field & \tableheadfont Description \\
\midrule
Use Case ID       & UC-05 \\
\rowcolor{tableRowAlt}Use Case Name     & Run Defect Detection Inference \\
Actor             & Quality Engineer \\
\rowcolor{tableRowAlt}Pre-condition     & Trained EMA checkpoint exists; input HPDC image is available \\
Description       & The quality engineer submits an HPDC image. The system slices it into 1024$\times$1024 patches, runs RF-DETR on each, and returns annotated detections. \\
\rowcolor{tableRowAlt}Basic Flow        & 1.\ Engineer provides image path. \newline 2.\ System generates patches (1024 px, 35\% overlap). \newline 3.\ RF-DETR processes each patch. \newline 4.\ Detections mapped to full-image coordinates. \newline 5.\ Annotated output image returned. \\
Alternate Flow    & No defects found in any patch $\to$ system returns ``No defects detected''. \\
\rowcolor{tableRowAlt}Exception Flow    & Image below minimum resolution $\to$ error raised. \\
Post-condition    & Annotated image saved with bounding boxes for each detected defect. \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Use Case UC-03: Train RF-DETR Model}
\begin{tabular}{p{3.5cm} p{9cm}}
\toprule
\rowcolor{tableHeader}\tableheadfont Field & \tableheadfont Description \\
\midrule
Use Case ID       & UC-03 \\
\rowcolor{tableRowAlt}Use Case Name     & Train RF-DETR Model \\
Actor             & ML Engineer \\
\rowcolor{tableRowAlt}Pre-condition     & Patch dataset generated and split in COCO format \\
Description       & ML engineer initiates training with configurable hyperparameters. \\
\rowcolor{tableRowAlt}Basic Flow        & 1.\ Engineer sets batch size, epochs, resolution. \newline 2.\ System loads DINOv2 pre-trained backbone. \newline 3.\ Training loop runs with EMA tracking. \newline 4.\ Best EMA checkpoint saved. \\
Alternate Flow    & VRAM insufficient $\to$ batch size reduced automatically. \\
\rowcolor{tableRowAlt}Exception Flow    & Loss becomes NaN $\to$ training halts, error logged. \\
Post-condition    & Best EMA checkpoint saved; mAP50-95 reported on validation set. \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Nonfunctional Requirements}

\begin{itemize}[leftmargin=2em]
    \item \textbf{NFR1 --- Performance:} Inference on a single 1024$\times$1024 patch shall complete in under 20~ms on a GPU (T4-class or better).
    \item \textbf{NFR2 --- Accuracy:} The system shall achieve a minimum EMA mAP50-95 of 0.20 on the test set to be considered production-viable for Phase 1.
    \item \textbf{NFR3 --- Licence Compliance:} All model weights, frameworks, and key libraries shall use Apache 2.0, MIT, or BSD licences compatible with commercial deployment.
    \item \textbf{NFR4 --- Reproducibility:} All experiments shall be reproducible from documented configurations (patch size, overlap, balancing ratio, augmentation, epochs, batch size).
    \item \textbf{NFR5 --- Scalability:} The patch generator shall handle any input resolution above 1024$\times$1024 without code modification.
    \item \textbf{NFR6 --- Maintainability:} The pipeline shall be modular (separate patch generation, training, and inference scripts) to allow independent updates.
    \item \textbf{NFR7 --- Security:} Model checkpoints and proprietary HPDC images shall remain under Magna International's data governance policy and not be distributed publicly.
    \item \textbf{NFR8 --- Usability:} The inference module shall produce annotated output images with bounding boxes overlaid, requiring no additional post-processing by the quality engineer.
\end{itemize}

\section{Software and Hardware requirement specifications}

\begin{table}[H]
\centering
\caption{Hardware Requirements}
\begin{tabular}{ll}
\toprule
\rowcolor{tableHeader}\tableheadfont Component & \tableheadfont Specification \\
\midrule
GPU (Training)  & NVIDIA RTX 6000 Ada (49 GB VRAM) or equivalent \\
\rowcolor{tableRowAlt}GPU (Inference) & NVIDIA T4 or better (16 GB VRAM minimum) \\
CPU             & 16-core or better (for data loading workers) \\
\rowcolor{tableRowAlt}RAM             & 64 GB minimum for patch generation and training \\
Storage         & 500 GB SSD (raw images, patches, checkpoints) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Software Requirements}
\begin{tabular}{ll}
\toprule
\rowcolor{tableHeader}\tableheadfont Software / Library & \tableheadfont Version / Details \\
\midrule
Operating System  & Ubuntu 22.04 LTS \\
\rowcolor{tableRowAlt}Python            & 3.10+ \\
PyTorch           & 2.0+ with CUDA 11.8 \\
\rowcolor{tableRowAlt}RF-DETR           & Roboflow RF-DETR (Apache 2.0) \\
Shapely           & 2.0+ (polygon clipping) \\
\rowcolor{tableRowAlt}OpenCV            & 4.8+ (image I/O) \\
pycocotools       & 2.0+ (COCO evaluation) \\
\rowcolor{tableRowAlt}NumPy             & 1.24+ \\
Matplotlib        & 3.7+ (visualisation) \\
\bottomrule
\end{tabular}
\end{table}
