\chapter{RESULTS AND DISCUSSIONS}

\justify

This chapter presents the results of all training experiments and a comparative analysis of performance improvements achieved through dataset engineering. The primary metric is COCO-standard EMA mAP50-95 (averaged over IoU thresholds 0.50--0.95).

\begin{table}[H]
\centering
\caption{Complete Experiment Comparison}
\begin{tabular}{lcccccc}
\toprule
\rowcolor{tableHeader}
{\color{white}\bfseries Exp} &
{\color{white}\bfseries Patch} &
{\color{white}\bfseries Overlap} &
{\color{white}\bfseries Balanced} &
{\color{white}\bfseries Aug} &
{\color{white}\bfseries Reg mAP} &
{\color{white}\bfseries EMA mAP} \\
\midrule
1 -- Baseline  & 768  & 20\% & No  & None   & 0.2006 & 0.2144 \\
\rowcolor{tableRowAlt}
4 -- Bal+Aug   & 768  & 20\% & Yes & H-Flip & 0.2077 & 0.2157 \\
7 -- 1024+35\% & 1024 & 35\% & No  & H-Flip & \textbf{0.2271} & \textbf{0.2379} \\
\bottomrule
\end{tabular}
\end{table}

\section*{Experiment 1 --- Baseline (768 px Patches)}

Training on 3,460 unbalanced patches (768$\times$768, 20\% overlap) with no augmentation established the reference level. EMA mAP50-95 reached \textbf{0.2144}. Key bottlenecks: 80.2\% of patches contained no defects (severe class imbalance), and small crack annotations were frequently truncated at patch boundaries due to low overlap.

\section*{Experiment 4 --- 768 px Balanced and Augmented}

Downsampling negatives to a 1:2 ratio and applying horizontal flip to crack patches improved Crack AP50-95 by \textbf{+17.4\%} (0.154 $\to$ 0.181) and overall EMA mAP to \textbf{0.2157}. This confirmed that dataset quality is the primary performance driver, not model architecture.

\section*{Experiment 7 --- 1024 px Patches with 35\% Overlap (Current Best)}

Increasing patch size to 1024$\times$1024 and overlap to 35\% provided each patch with more visual context and reduced boundary truncation. The positive patch ratio improved naturally to 31.98\% without explicit balancing.

\begin{resultbox}[Experiment 7 --- Best Results]
\begin{tabular}{ll}
Best EMA mAP50-95:     & \textbf{0.2379} \\
Best Regular mAP50-95: & 0.2271 \\
Best mAP50:            & \textbf{0.4647} \\
Peak performance:      & Epoch 8--10 \\
Checkpoint:            & \texttt{checkpoint\_best\_ema.pth} \\
\end{tabular}
\end{resultbox}

\section*{Training Curve Summary (Experiment 7)}

\begin{table}[H]
\centering
\caption{EMA mAP50-95 at Key Training Epochs (Experiment 7)}
\begin{tabular}{ccccccc}
\toprule
\rowcolor{tableHeader}
{\color{white}\bfseries Epoch} &
{\color{white}\bfseries 1} &
{\color{white}\bfseries 5} &
{\color{white}\bfseries 8} &
{\color{white}\bfseries 10 (peak)} &
{\color{white}\bfseries 15} &
{\color{white}\bfseries 20} \\
\midrule
EMA mAP50-95 & 0.170 & 0.218 & 0.235 & \textbf{0.2379} & 0.227 & 0.220 \\
\bottomrule
\end{tabular}
\end{table}

Performance peaks at Epoch 8--10 and then mildly declines, indicating the dataset size limits generalisation beyond this point. The EMA checkpoint saved at peak epoch is recommended for deployment.

\section*{Key Findings}

\begin{enumerate}[leftmargin=2em]
    \item \textbf{Dataset engineering is the dominant factor.} All three training runs used identical RF-DETR Medium architecture. The full 11\% improvement came entirely from data pipeline changes (patch size, overlap, balancing, augmentation).
    \item \textbf{Patch size is the single most impactful variable.} Scaling from 768 to 1024 px with 35\% overlap produced a \textbf{+10.3\%} relative improvement in EMA mAP50-95, the largest single gain observed.
    \item \textbf{Crack-specific augmentation is effective.} Horizontal flip on crack patches alone gave \textbf{+17.4\%} Crack AP in Experiment 4 vs.\ Experiment 1.
    \item \textbf{EMA checkpoint is consistently superior.} In all three experiments the EMA checkpoint outperformed the regular checkpoint, confirming that EMA tracking is essential for fine-tuning on small datasets.
\end{enumerate}

\chapter{CONCLUSIONS AND FUTURE SCOPE}

\section*{Conclusions}

This project successfully designed, implemented, and validated an automated deep learning pipeline for surface defect detection in HPDC aluminium castings, developed in partnership with Magna International. The system addresses a real industrial need: existing EOL inspection methods (human visual inspection and periodic X-Ray) are either unreliable or too slow for 100\% production coverage.

The following outcomes were achieved:

\begin{itemize}[leftmargin=2em]
    \item \textbf{RF-DETR adopted as a commercially compliant model.} YOLOv8 and YOLO11, initially evaluated, were excluded due to their AGPL-3.0 licence incompatibility with Magna's proprietary production systems. RF-DETR (Apache 2.0) was validated as an equivalent-performance alternative with no commercial deployment restrictions.
    \item \textbf{Patch-based pipeline for extreme small-object detection.} A sliding-window generator converts 256 full HPDC images into 4,184 training patches (1024$\times$1024, 35\% overlap) with Shapely polygon-aware annotation clipping, enabling detection of cracks occupying less than 0.1\% of image area.
    \item \textbf{EMA mAP50-95 of 0.2379 achieved.} This represents an approximately 11\% improvement over the unoptimised baseline of 0.2144, achieved entirely through systematic dataset engineering.
    \item \textbf{Patch size identified as the dominant variable.} Scaling from 768 to 1024 px with 35\% overlap produced the largest single performance gain across all experiments.
\end{itemize}

\begin{resultbox}[Final Best Configuration]
\begin{tabular}{ll}
Model:          & RF-DETR Medium (Apache 2.0) \\
Patch size:     & 1024 $\times$ 1024 px, 35\% overlap \\
Augmentation:   & Horizontal Flip (crack patches only) \\
EMA mAP50-95:   & \textbf{0.2379} \quad|\quad mAP50: \textbf{0.4647} \\
Checkpoint:     & \texttt{checkpoint\_best\_ema.pth} \\
\end{tabular}
\end{resultbox}

\section*{Future Scope}

\begin{itemize}[leftmargin=2em]
    \item \textbf{RF-DETR Large:} Evaluate the larger DINOv2 ViT-L backbone on the same 1024 dataset. Expected gain: $+$0.01--0.03 mAP50-95.
    \item \textbf{Extended augmentation:} Add vertical flip, 90°/180° rotation, and brightness jitter to improve generalisation and delay overfitting beyond Epoch 10.
    \item \textbf{Increased overlap (50\%):} Further reduce boundary truncation for very long crack annotations.
    \item \textbf{Phase 2 --- Crack Length Measurement:} Extend the pipeline to pixel-level segmentation, Zhang-Suen morphological skeletonisation, graph-based centerline extraction, and crack length estimation in physical units (mm) using a pixel-to-mm calibration factor.
    \item \textbf{Production integration:} Deploy the trained inference module on Magna's EOL inspection station with a real-time defect reporting dashboard, enabling 100\% automated surface inspection coverage.
\end{itemize}
