\chapter{PROPOSED SYSTEM}

\justify

The proposed system is an end-to-end deep learning pipeline for automated surface defect detection in HPDC aluminium castings. It addresses the fundamental challenge of detecting extremely small defect regions (crack area $<$ 0.1\% of image) in high-resolution industrial images, using a commercially deployable detection transformer.

\section{Description of Proposed System.}
\justify

The system operates as a two-stage pipeline. First, a sliding-window patch generator divides full 4096$\times$3000 HPDC images into 1024$\times$1024 overlapping patches (35\% overlap) with Shapely polygon-aware annotation clipping. Second, RF-DETR Medium is fine-tuned on this patch dataset and produces bounding box predictions with class labels and confidence scores, end-to-end without NMS.

\begin{figure}[H]
\centering
\begin{tikzpicture}[node distance=0.6cm,
block/.style={rectangle, rounded corners=4pt, draw=magnaDark,
fill=magnaLightGray, text width=5.5cm,
minimum height=1cm, align=center, font=\small},
arrow/.style={-Stealth, thick, color=magnaRed}]
\node[block, fill=magnaRed!15, draw=magnaRed] (raw)
    {\textbf{Raw HPDC Image} --- 4096$\times$3000 px};
\node[block, below=of raw] (patch)
    {\textbf{Patch Generator}\\Sliding window, Shapely clipping\\1024$\times$1024 px, 35\% overlap};
\node[block, below=of patch] (ds)
    {\textbf{Dataset Engineering}\\Negative balancing + Crack augmentation};
\node[block, below=of ds] (train)
    {\textbf{RF-DETR Medium Training}\\DINOv2 backbone + Deformable decoder\\EMA checkpoint tracking};
\node[block, fill=successGreen!15, draw=successGreen, below=of train] (infer)
    {\textbf{Inference Output}\\Bounding boxes: Crack / Cold Shut\\No NMS required};
\draw[arrow] (raw)   -- (patch);
\draw[arrow] (patch) -- (ds);
\draw[arrow] (ds)    -- (train);
\draw[arrow] (train) -- (infer);
\end{tikzpicture}
\caption{High-level pipeline of the proposed HPDC defect detection system.}
\label{fig:system_overview}
\end{figure}

The Figure~\ref{fig:system_overview} illustrates the complete flow from raw casting image to bounding box output. The patch-based strategy is the central design decision enabling small-object crack detection.

\section{Description of Target Users }
\justify

\begin{itemize}[leftmargin=2em]
    \item \textbf{Quality Control Engineers (Magna International):} End users who run inference on new casting images at the production line. They receive annotated images with bounding boxes via a command-line or GUI interface.
    \item \textbf{ML Engineers / Research Team:} Users who operate the data engineering pipeline and training framework, configure hyperparameters, and monitor mAP metrics to iterate on model performance.
\end{itemize}

\section{Advantages/Applications  of Proposed System}
\justify

\begin{itemize}[leftmargin=2em]
    \item \textbf{Automated and consistent inspection:} Removes inspector fatigue and variability. Every casting is evaluated reproducibly at a fixed confidence threshold.
    \item \textbf{Extreme small-object detection:} The 1024$\times$1024 patch strategy with 35\% overlap is specifically engineered for cracks occupying $<$ 0.1\% of image area.
    \item \textbf{Commercial deployment ready:} RF-DETR's Apache 2.0 licence permits unrestricted commercial use and deployment.
    \item \textbf{End-to-end without NMS:} Detections are produced via bipartite matching, eliminating the need to tune non-maximum suppression thresholds.
    \item \textbf{Extensible to Phase 2:} The bounding box output is a direct input to the planned Phase 2 segmentation and crack length measurement pipeline.
\end{itemize}

\section{Scope (Boundary of proposed system)}
\justify

\textbf{In scope (Phase 1):}
\begin{itemize}[leftmargin=2em]
    \item Detection and classification of Crack and Cold Shut defects via bounding boxes.
    \item Patch-based processing of HPDC images up to 4096$\times$3000 resolution.
    \item Training, evaluation, and EMA checkpoint management for RF-DETR Medium.
    \item COCO-standard mAP evaluation on train/val/test splits.
\end{itemize}

\textbf{Out of scope (Phase 2, deferred):}
\begin{itemize}[leftmargin=2em]
    \item Pixel-level instance segmentation and morphological skeletonisation.
    \item Crack length estimation in physical units (mm).
    \item Integration with Magna's production line control systems.
    \item Real-time video stream processing.
\end{itemize}
